A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-03-30 paper/paper.tex

\section{Known Neurobiological Properties of the Default Mode Network}
We begin by a deconstruction the overall DMN function
by the mechanistic relevance of each individual network node
based on published findings from neuroscience experiments.
%
\subsection{The posteromedial cortex: Global monitoring and integration}
The posteriormedial cortex (PMC) and
the dorsomedial prefrontal cortex (dmPFC) within the DMN
are responsible for the highest metabolic turn-over
of glucose energy consumption in humans \citep{raichle2001pnas}.
The PMC myelinates relatively late during postnatal development in monkeys
\citep{goldman1987development}, generally considered to
be a sign of phylogenetic sophistication (Flechsig, 1920).
Physiological and disturbed metabolic fluctuations in the
human PMC have been repeatedly related to
phenomena of changed conscious awareness,
including anesthesia,
sleep, and forms of coma \citep{cavanna2006precuneus}.
%
The PMC has long been speculated to reflect constant computation of
the environmental statistics and its internal representation
as an "inner minds eye" \citep{cavanna2006precuneus}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the bilateral parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to bind various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in the gathering
of information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity modalities
with the amygdala involved in significance evaluation and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the human ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.
%
Indeed,
electrophysiological recordings in animals implicated the PMC in
strategic selection \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
and outcome-contingent behavioral modulation \citep{hayden2009electrophysiological},
while its retrosplenial portion was
more specifically implicated in approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy in a complex food foraging task \citep{pearson2009neurons}.
Single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.

@@

\textbf{XXX Your focus in this subsection ought to be on the SC itself not POMPDs, etc.}
 \begin{itemize}
    \item Processing social information has been proposed to underlie the
    physiological baseline of human brain function
    \citep{schilbach2008minds}. This was later challenged by observing
    analogues of the DMN in monkeys \citep{mantini2011default}
    and rats \citep{lu2012rat}, with
    supposedly less advanced social capacities.
    The MDPs parsimoneously explain the dominance of social content in
    human mental scenes by their extremely high relevance to humans
    reflected by high values for social information in the value matrix
    \citep{baker2009action}.
    \item How can a same neurobiological circuit be equally important
    for baseline house-keeping functions and specific task performance?
    Our computational account can explain why the DMN is implicated
    in both a goal-directed task and an idlying rest cognitive set,
    if environmental relevance is processed, manipulated, and retained
    also as a baseline function of human mental activity.
    In tasks, the policy and value matrices are updated to optimize short-term action,
    whereas, at rest, these parameter updates may
    improve especially mid- and long-term action.
    \item The sentinel account does not provide a formal mechanisms of
    how attention and memory resources are exactly reallocated when
    encountering a salient environmental stimulus. \textbf{XXX remove the next sentence. Danilo: can you help me to correct the sentence instead.
    There appears to be a cool link between cognition and formal theory
    here.}
    In contrast,
    the Bellman recursion is weighted at each step by
    $\sum_{a \in A}\pi(a|s)$. This factor weighs which action cascades
    are recursively explored by the agent and which decision trees are neglected.
    This directly implies allocation of attention and computational load.
    \item In sum,
    MDPs imply a ``radar'' function of monitoring the environment for salient information
    in general and explain social thinking as a physiological brain baseline
    as an important special case, without excluding other topics
    contemplated in continuous mental activity.
 \end{itemize}

>>>>

\section{Known Neurobiological Properties of the Default Mode Network}
\textbf{XXX (danilo) this section needs more tight integration of the parts, and no just a separate independent subsection for each portion of the DMN...}
We begin by a deconstruction the overall DMN function
by the mechanistic relevance of each individual network node
based on published findings from neuroscience experiments.
%
\subsection{The posteromedial cortex: Global monitoring and integration}
The posteriormedial cortex (PMC) and
the dorsomedial prefrontal cortex (dmPFC) within the DMN
are responsible for the highest metabolic turn-over
of glucose energy consumption in humans \citep{raichle2001pnas}.
The PMC myelinates relatively late during postnatal development in monkeys
\citep{goldman1987development}, generally considered to
be a sign of phylogenetic sophistication (Flechsig, 1920).
Physiological and disturbed metabolic fluctuations in the
human PMC have been repeatedly related to
phenomena of changed conscious awareness,
including anesthesia,
sleep, and forms of coma \citep{cavanna2006precuneus}.
%
The PMC has long been speculated to reflect constant computation of
the environmental statistics and its internal representation
as an "inner minds eye" \citep{cavanna2006precuneus}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the bilateral parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to bind various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in the gathering
of information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity modalities
with the amygdala involved in significance evaluation and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the human ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.
%
Indeed,
electrophysiological recordings in animals implicated the PMC in
strategic selection \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
and outcome-contingent behavioral modulation \citep{hayden2009electrophysiological},
while its retrosplenial portion was
more specifically implicated in approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy in a complex food foraging task \citep{pearson2009neurons}.
Single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.

@@

\textbf{XXX (danilo) Your focus in this subsection ought to be on the SC itself not POMPDs, etc.}
 \begin{itemize}
    \item Processing social information has been proposed to underlie the
    physiological baseline of human brain function
    \citep{schilbach2008minds}. This was later challenged by observing
    analogues of the DMN in monkeys \citep{mantini2011default}
    and rats \citep{lu2012rat}, with
    supposedly less advanced social capacities.
    The MDPs parsimoneously explain the dominance of social content in
    human mental scenes by their extremely high relevance to humans
    reflected by high values for social information in the value matrix
    \citep{baker2009action}.
    \item How can a same neurobiological circuit be equally important
    for baseline house-keeping functions and specific task performance?
    Our computational account can explain why the DMN is implicated
    in both a goal-directed task and an idlying rest cognitive set,
    if environmental relevance is processed, manipulated, and retained
    also as a baseline function of human mental activity.
    In tasks, the policy and value matrices are updated to optimize short-term action,
    whereas, at rest, these parameter updates may
    improve especially mid- and long-term action.
    \item The sentinel account does not provide a formal mechanisms of
    how attention and memory resources are exactly reallocated when
    encountering a salient environmental stimulus. \textbf{XXX remove the next sentence. Danilo: can you help me to correct the sentence instead.
    There appears to be a cool link between cognition and formal theory
    here.}
    In contrast,
    the Bellman recursion is weighted at each step by
    $\sum_{a \in A}\pi(a|s)$. This factor weighs which action cascades
    are recursively explored by the agent and which decision trees are neglected.
    This directly implies allocation of attention and computational load.
    \item In sum,
    MDPs imply a ``radar'' function of monitoring the environment for salient information
    in general and explain social thinking as a physiological brain baseline
    as an important special case, without excluding other topics
    contemplated in continuous mental activity.
 \end{itemize}
